---
title: "getting-started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(terra)
library(sf)
library(catchment)
```

## Loading data

To get started with the catchment package, we first need to load the required data. At a minimum, we will need locations for our health facilities, an initial weight for each health facility to use as a proxy for attractiveness, and a boundary to determine the extent. We will also need a raster of the population that we want to distribute.

We have provided some data within the package for this tutorial to get started. The boundary (`example_shp`, an `sf` object) and locations (`example_locs`, a data frame) are lazy-loaded datasets, while the population raster is loaded with the `example_pop()` function (a [`terra::SpatRaster`] cannot be stored as a lazy-loaded dataset).

```{r load-data}
data("example_shp")
data("example_locs")
pop <- example_pop()
```

Next we will also need a friction surface, which we will use to calculate travel times from each pixel to all health facility locations. In this tutorial, we can use the friction surface developed by the Malaria Atlas Project. Our team has written a function for pulling this surface in a package called [PATHtools](https://github.com/PATH-Global-Health/PATHtools), which you can install via GitHub.

We also need to resample the friction raster so that it aligns with our population raster.

```{r fric-surface}
# devtools::install_github("PATH-Global-Health/PATHtools")
fric <- terra::rast(PATHtools::get_friction_surface(example_shp))
fric <- terra::resample(fric, pop, method = "average")
```

Finally, we need to designate an output location. For this tutorial we will use a temporary folder.

```{r}
f <- tempfile()
```


## Travel time surfaces

Next, we need to create travel time rasters for each health facility using the friction surface. This can be a time consuming process depending on the number of health facilities, size of the boundary file, and resolution of our population raster. Therefore, we prefer to save the rasters locally so that they only need to be rendered once.

`create_travel_surface()` uses [`terra::costDist()`] to compute accumulated cost-distance (travel time) from each facility.

```{r}
# Create a folder to save travel time (tt) rasters
fs::dir_create(fs::path(f, "tt"))

# Generate travel time rasters
create_travel_surface(
  friction_surface = fric,
  extent_file = pop,
  points = example_locs,
  id_col = "label",
  x_col = "x",
  y_col = "y",
  output_dir = fs::path(f, "tt"),
  individual_surfaces = TRUE)
```

Next, we need to organize the travel time values into a matrix, where each column is a health facility and each row is a pixel from our population raster. To reduce the computational burden, we only want to keep pixels that have population values; this helps make the matrix sparse and improves model fitting.

We can use `travel_mat_from_folder()` to generate this matrix. Note that this function will pull in values for all `.tif` files in the reference folder, so it is important to keep the travel time rasters in their own folder. Also, health facilities will be organized in alphanumeric order based on the file names (the `id_col` values). This ensures that we maintain alignment between files.

```{r}
## Organize travel time matrix
tmat <- travel_mat_from_folder(dir = fs::path(f, "tt"),
                               reference = pop)
```

## Preparing the model data

Before fitting the catchment model, we organize our data using the `prepare_data()` function. This function standardizes the input data and generates the INLA mesh.

> Note: fitting the model requires the [INLA](https://www.r-inla.org/download-install) and [TMB](https://github.com/kaskr/adcomp) packages. See the README for installation instructions.

There are two ways to supply the accessibility information, depending on whether you want the model to **learn** the distance-decay shape or use a **fixed** one.

**Recommended — learn the decay.** Pass the raw travel-time matrix (`tmat`) directly. `prepare_data()` then builds the sparse travel template the model needs to estimate the decay parameter. The sparsity controls live here: `minimum_time` clamps very short travel times (the power family blows up as travel time approaches zero), `force_threshold` forces remote pixels to their nearest facility, and `n_fac_limit` caps how many facilities each pixel can be allocated to.

```{r}
## Organize input data (learned-decay path)
catch_dat <- prepare_data(
  prob_mat_init  = tmat,
  pop_raster     = pop,
  location_data  = example_locs,
  minimum_time   = 10,
  force_threshold = 300,
  n_fac_limit    = 10,
  mesh.args = list(cutoff = 0.1, max.edge = c(0.1, 4)))
```

**Alternative — fixed decay.** If you prefer to bake in a fixed decay shape (e.g. inverse-squared distance), build the probability matrix yourself with `initial_access_surface()` and pass that instead. Fitting then uses `decay = "none"` (see below).

```{r}
## Create a fixed initial accessibility matrix (legacy path)
pmat <- initial_access_surface(
  tmat,
  n_fac_limit = 10,
  force_threshold = 300,
  sparse = FALSE)

catch_dat_fixed <- prepare_data(
  prob_mat_init = pmat,
  pop_raster    = pop,
  location_data = example_locs,
  mesh.args = list(cutoff = 0.1, max.edge = c(0.1, 4)))
```

Before fitting the model, we should first check the INLA mesh to make sure that it is not too dense or sparse. These parameters can be adjusted using the `mesh.args` argument.

```{r}
## View INLA mesh
plot(catch_dat$mesh)
plot(sf::st_geometry(example_shp), add = TRUE, border = "red", lwd = 2)
points(example_locs$x, example_locs$y)
```

Now we can fit the catchment model. Please keep in mind that this is a computationally intensive model that will require significant RAM depending on the area of interest, number of health facilities, and population resolution.

```{r message=FALSE, warning=FALSE, results='hide'}
## Fit catchment model
mod <- catchment_model(catch_dat)
```

By default `catchment_model()` fits a Poisson likelihood and **learns an exponential distance-decay** (`exp(-d / tau)`, with the scale `tau` in minutes estimated from the data). You can change either of these:

```{r}
## Power-law decay (estimate the exponent a in d^-a)
mod_pow <- catchment_model(catch_dat, decay = "power")

## Negative-binomial likelihood for overdispersed counts
mod_nb  <- catchment_model(catch_dat, family = "nb")

## Fixed decay (uses the precomputed surface in catch_dat_fixed)
mod_fix <- catchment_model(catch_dat_fixed, decay = "none")
```

The learned decay parameter is stored on the fitted object:

```{r}
mod$decay        # "exponential"
mod$decay_param  # estimated tau (minutes)
```

## Checking convergence

`catchment_model()` checks convergence automatically and warns if the optimiser failed, the gradient is large, or the Hessian is not positive-definite. You can inspect the diagnostics at any time:

```{r}
check_convergence(mod)
mod                 # print method shows family, decay, convergence, pdHess
```

## Post-processing results

Once the model is fit, we can use the `catchment_populations()` function to generate the estimated catchment population for each facility.

```{r}
## Estimated catchment populations (point estimates)
catchment_populations(mod)
example_locs$est_pop <- catchment_populations(mod)
```

Because the model is fit with TMB, standard errors are propagated through the random effects via `TMB::sdreport()`. Set `uncertainty = TRUE` to return a data frame with standard errors and approximate 95% confidence intervals:

```{r}
## Estimated catchment populations with uncertainty
catchment_populations(mod, uncertainty = TRUE)
```

## Validating the fit

The package provides posterior-predictive checks and leave-one-facility-out cross-validation to assess how well the model reproduces the observed facility counts.

```{r}
## Posterior-predictive check: coverage + dispersion + a diagnostic plot
ppc <- pp_check(mod, nsim = 1000)
ppc$coverage     # fraction of observed counts inside their predictive interval
ppc$dispersion   # ~1 indicates Poisson-consistent dispersion
ppc$plot         # ggplot of observed counts vs predictive intervals
```

```{r}
## Leave-one-facility-out cross-validation
cv <- loo_facility_cv(mod)
attr(cv, "rmse")
attr(cv, "mae")
```

## Visualizing catchments

We can map each pixel to the facility it is most likely to attend with `dominant_catchment()`, or plot the access-probability surface for a single facility with `plot_prob_surface()`. The plot methods use **ggplot2** and **tidyterra** when available (and fall back to base `terra` plotting otherwise).

```{r}
## Dominant-catchment map (also the default plot() for a fitted model)
plot(mod)

## Access-probability surface for one facility
plot_prob_surface(mod, id_label = example_locs$label[1])

## Or extract the raster directly
prob_r <- get_prob_raster(mod, id_label = example_locs$label[1])
```
