---
title: "Prior-predictive checks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prior-predictive checks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
set.seed(1)
```

The catchment model places priors on several quantities. Before fitting it is
worth understanding what those priors *imply* about the model's behaviour — this
is the idea behind a **prior-predictive check**. This vignette visualises the
priors the package uses and shows how sensitive the implied catchment allocation
is to the two priors you are most likely to want to tune: the facility-mass
prior and the distance-decay prior.

None of the chunks below fit a model, so this vignette runs without INLA or TMB.

## The facility-mass prior

Each facility carries a multiplicative *mass* (attractiveness) term. The model
works on the log scale with

$$\texttt{log\_hf\_mass} \sim \mathrm{Normal}(0,\ \sigma_\text{mass}),$$

so on the natural scale the mass is log-normal and centred on 1. The default
`log_hf_mass_sd = 0.1` is deliberately tight: it keeps facilities close to equal
attractiveness and stabilises a model that is often weakly identified when there
are few facilities. Loosening it lets the data pull facility masses further
apart.

The plot below shows the implied prior on the *relative* mass (mass divided by
its median) for three values of the prior SD.

```{r mass-prior}
sds <- c(0.1, 0.5, 1.0)
x   <- seq(0.3, 3, length.out = 400)         # relative mass
cols <- c("black", "steelblue", "firebrick")

plot(NA, xlim = range(x), ylim = c(0, 4.2),
     xlab = "relative facility mass", ylab = "prior density",
     main = "Implied prior on relative facility mass")
for (i in seq_along(sds)) {
  lines(x, dlnorm(x, meanlog = 0, sdlog = sds[i]), col = cols[i], lwd = 2)
}
abline(v = 1, lty = 3, col = "grey60")
legend("topright", legend = paste0("sd = ", sds),
       col = cols, lwd = 2, bty = "n")
```

A useful summary is the 95% prior interval for the mass *ratio* between two
facilities — how different the model is willing to let two facilities' pulls be,
a priori:

```{r mass-ratio}
ratio_interval <- function(sd) {
  # ratio of two independent log-normal masses is log-normal with sd*sqrt(2)
  round(exp(qnorm(c(0.025, 0.975)) * sd * sqrt(2)), 2)
}
sapply(sds, ratio_interval)
```

With the default `sd = 0.1` two facilities differ by at most ~30% a priori;
at `sd = 1` one facility can plausibly attract several times the population of
another before seeing any data.

## The distance-decay prior

When the decay parameter is learned (the default), the model estimates it on the
log scale with a `Normal(log(centre), 1)` prior, where the centre is the family
default: `tau = 60` minutes for the exponential family and `a = 2` for the power
family. The wide SD of 1 on the log scale is intentionally weakly informative.

First, the prior on the decay parameter itself:

```{r decay-param-prior}
tau   <- seq(5, 300, length.out = 400)
a     <- seq(0.3, 6, length.out = 400)

par(mfrow = c(1, 2))
plot(tau, dlnorm(tau, meanlog = log(60), sdlog = 1), type = "l", lwd = 2,
     xlab = expression(tau ~ "(minutes)"), ylab = "prior density",
     main = "Exponential scale")
abline(v = 60, lty = 3, col = "grey60")

plot(a, dlnorm(a, meanlog = log(2), sdlog = 1), type = "l", lwd = 2,
     xlab = "exponent a", ylab = "prior density",
     main = "Power exponent")
abline(v = 2, lty = 3, col = "grey60")
par(mfrow = c(1, 1))
```

More informative is the *prior-predictive* spread of the decay **curves** these
priors imply. We draw decay parameters from the prior and plot the resulting
weight-vs-travel-time relationships. Steeper curves mean access falls off
quickly with travel time (tight, local catchments); flatter curves mean facilities
draw from further away.

```{r decay-curves}
d <- seq(1, 180, length.out = 200)          # travel time (minutes)
n_draw <- 60

par(mfrow = c(1, 2))

## Exponential: exp(-d / tau)
plot(NA, xlim = range(d), ylim = c(0, 1),
     xlab = "travel time (min)", ylab = "relative access weight",
     main = "Exponential decay (prior draws)")
tau_draws <- rlnorm(n_draw, meanlog = log(60), sdlog = 1)
for (t in tau_draws) lines(d, exp(-d / t), col = adjustcolor("steelblue", 0.3))
lines(d, exp(-d / 60), col = "black", lwd = 2)   # prior median

## Power: d^-a, normalised to 1 at d = 1 for display
plot(NA, xlim = range(d), ylim = c(0, 1),
     xlab = "travel time (min)", ylab = "relative access weight",
     main = "Power decay (prior draws)")
a_draws <- rlnorm(n_draw, meanlog = log(2), sdlog = 1)
for (aa in a_draws) lines(d, d^(-aa) / 1^(-aa), col = adjustcolor("firebrick", 0.3))
lines(d, d^(-2), col = "black", lwd = 2)         # prior median
par(mfrow = c(1, 1))
```

The exponential family puts most of its prior mass on smooth, moderate decay,
while the power family admits both very steep (near-nearest-facility) and very
flat allocations. If you have strong external knowledge about how far people
travel, you can fix the decay instead of learning it (`estimate_decay = FALSE`,
`decay_init = ...`).

## The spatial-field priors

The Matérn SPDE field that captures residual spatial structure in demand has
priors on its range (`log_rho`) and marginal standard deviation (`log_sigma`).
The defaults are `log_rho ~ Normal(log(5), 0.5)` and
`log_sigma ~ Normal(-1, 0.5)`.

```{r field-priors}
par(mfrow = c(1, 2))
rho   <- seq(0.5, 30, length.out = 400)
sigma <- seq(0.05, 2, length.out = 400)

plot(rho, dlnorm(rho, meanlog = log(5), sdlog = 0.5), type = "l", lwd = 2,
     xlab = "range (spatial units)", ylab = "prior density",
     main = "Field range")
plot(sigma, dlnorm(sigma, meanlog = -1, sdlog = 0.5), type = "l", lwd = 2,
     xlab = "marginal SD", ylab = "prior density",
     main = "Field marginal SD")
par(mfrow = c(1, 1))
```

The range prior should be interpreted relative to the extent of your study area
and the units of your coordinates. If your domain is much larger or smaller than
these defaults assume, adjust the mesh (via `mesh.args` in `prepare_data()`) and
consider whether the range prior is sensible for your spatial scale.

## Putting it together

A practical prior-predictive workflow is:

1. Decide how different you believe facility attractiveness can be, and set
   `log_hf_mass_sd` so the implied mass-ratio interval matches that belief.
2. Decide whether access decay is local or broad, choose the decay family, and
   either learn it (default) or fix it if you have strong prior knowledge.
3. Check that the field range prior is reasonable for your spatial extent.

Once fitted, validate the model against the observed counts with `pp_check()`
and `loo_facility_cv()` — see the *getting-started* vignette.
